\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2018-01-12 eg6-letters/NIM/EG6_RTPC_NIM.tex
review of readout session

\begin{figure*}[tb]
   \centering
   \includegraphics[width=.75\textwidth]{ReadoutScheme.png}
   \caption[]{Schematic representation of the RTPC readout system, showing
   the different elements of the front end electronics.} 
   \label{fig:ReadoutScheme}
\end{figure*}

The readout system, similar to the original BoNuS RTPC system~\cite{Fenker:2008zz}, 
was based on the front end electronic boards originally developed for the ALICE 
TPC readout system at CERN~\cite{ALICE-FEE}. The Front End Card (FEC) hosted 
the analogue receiver circuit and the subsequent amplification and digitization 
stages, based on the PASA~\cite{Soltveit:2012jp} and 
ALTRO~\cite{EsteveBosch:2003bj} ASICS. A Readout Control Unit (RCU) board was 
used to distribute the trigger signal to FECs and for data readout, with each 
RCU handling up to 25 FECs. Communication between readout controller 
application (ROC) and FECs was performed trough a custom backplane, 
implementing a low-voltage signal bus. The RCU communicated with the CLAS DAQ 
system trough a 200 MB/s optical link, connected to a data acquisition PC 
hosting a readout receiver card. This PC hosted the ROC used to interface with 
the CLAS DAQ system. An Ethernet link was also present, for slow-controls and 
monitoring.

The standard CLAS electron trigger was used to initiate the RTPC readout. For each 
event, 100 samples/channel were digitized and processed by the ALTRO ASIC. In order 
to reduce the data size, ALTRO was operated in zero-suppression mode, keeping only 
samples from channels above a programmable threshold set just above the noise level. 
Samples were pre-processed 
by a glitch-filter to reject noise and spurious pulses. To properly reconstruct the 
signal shape, $N_{PRE}=3$ samples before threshold-crossing and $N_{POST}=3$ samples 
after the signal returns below threshold were saved.

The trigger signal also initiates RCU readout operation from FEC boards. All the 
measured samples from active channels are reported, together with a channel 
identifier and a timestamp, to the ROC application, that in turn sends them on 
the main event builder. Readout from FECs occurs in ``block-transfer'' mode, 
making use of the ALTRO internal multi-event buffer. This feature could be 
exploited thanks to the new RCU boards developed for the RTPC readout 
system\footnote{In the original BoNuS system, the U2F readout controller had no 
   multi-event capability and used a USB connection rather than fiber optics 
   for readout. These factors limited the maximum readout rate from the BoNuS 
detector to about 500 Hz.}. In order to read all the detector readout pads, 
four FEC crates were used, each equipped with 6 boards, plus a ROC. A schematic 
of the readout system, for a single crate, is reported in 
Figure~\ref{fig:ReadoutScheme}. This configuration permitted to reduce the dead 
time associated to FEC readout operations, which is scaling linearly with the 
number of boards in the crate. During the 2009 run, the system was successfully 
operated with a DAQ rate of 3.1 kHz and a live time of $70 \%$, for a 
luminosity of about $10^{34}$~cm$^{-2}\cdot$s$^{-1}$ and a beam energy of 
6.064~GeV.

Finally, during data reconstruction, the acquired samples were processed to obtain, 
for each readout pad, the accumulated charge (ADC) and the pulse time (T). Since 
pulse time was obtained as the time-stamp of the first sample above threshold, 
referred to the trigger time, the resolution is equivalent to the ALTRO 
sampling time, 100~ns.

@@

 \bibitem{ALICE-FEE}
 L. Musa et al., The ALICE TPC front end Electronics Nuclear Science Symposium 
 Conference Record, 2003 IEEE, 5 3647-3651 Vol.5, Oct (2003).

\bibitem{Soltveit:2012jp} 
   H.~K.~Soltveit {\it et al.},
  The Preamplifier shaper for the ALICE TPC-Detector,
  Nucl.\ Instrum.\ Meth.\ A {\bf 676}, 106 (2012).
  %doi:10.1016/j.nima.2012.02.012
  %[arXiv:1203.3564 [physics.ins-det]].

\bibitem{EsteveBosch:2003bj} 
   R.~Esteve Bosch, A.~Jimenez de Parga, B.~Mota and L.~Musa,
  The ALTRO chip: A 16-channel A/D converter and digital processor for gas 
  detectors,
  IEEE Trans.\ Nucl.\ Sci.\  {\bf 50}, 2460 (2003).
  %doi:10.1109/TNS.2003.820629

>>>>

%\begin{figure*}[tb]
%   \centering
%   \includegraphics[width=.75\textwidth]{ReadoutScheme.png}
%   \caption[]{Schematic representation of the RTPC readout system, showing
%   the different elements of the front end electronics.} 
%   \label{fig:ReadoutScheme}
%\end{figure*}

The readout system was an upgraded version of the original BoNuS RTPC system~\cite{Fenker:2008zz}, 
based ALICE-TPC front end electronic boards (FECs)~\cite{ALICE-FEE}. 
Each FEC hosted 128 channels, providing amplification and digitization of the input signals.  For each event, 100 samples/channel were digitized by the ALTRO ASIC~\cite{EsteveBosch:2003bj}, operated at 10 MHz sampling frequency. In order to reduce the data size, the system was operated in zero-suppression mode, keeping data from $N_{PRE}=3$ samples before threshold crossing on the rising edge to $N_{POST}=3$ samples after threshold crossing on the falling edge. The theshold level was set on each channel just above the noise level.

A new custom backplane was developed to connect FECs to the Readout Control Unit (RCU)~\cite{RCU}, allowing fast (200 MB/s) communication between the boards. The RCU board, used to distribute the trigger signal to different FECs and read data, was equipped with a fast optical link for data transferring to the main event builder. These features, together with the ``block-transfer'' readout mode making use of the FECs multi-event buffer, allowed to reach a signifantly higher readout-rate compared to the original BoNuS system.  During the 2009 run, the system was successfully 
operated with a DAQ rate of 3.1 kHz and a live time of $70 \%$, for a 
luminosity of about $10^{34}$~cm$^{-2}\cdot$s$^{-1}$ and a beam energy of 
6.064~GeV, to be compared with the 500 Hz obtained with the first BoNuS detector at similar run conditions.

During data reconstruction, the acquired samples were processed to obtain, 
for each readout pad, the accumulated charge (ADC) and the pulse time (T). Since 
pulse time was obtained as the time-stamp of the first sample above threshold, 
referred to the trigger time, the resolution was equivalent to the ALTRO 
sampling time of 100~ns.

@@

 \bibitem{ALICE-FEE}
 L. Musa et al., The ALICE TPC front end Electronics, Nuclear Science Symposium 
 Conference Record, 2003 IEEE, 5 3647-3651 Vol.5, Oct (2003).

%\bibitem{Soltveit:2012jp} 
%   H.~K.~Soltveit {\it et al.},
%  The Preamplifier shaper for the ALICE TPC-Detector,
%  Nucl.\ Instrum.\ Meth.\ A {\bf 676}, 106 (2012).
  %doi:10.1016/j.nima.2012.02.012
  %[arXiv:1203.3564 [physics.ins-det]].

\bibitem{EsteveBosch:2003bj} 
   R.~Esteve Bosch, A.~Jimenez de Parga, B.~Mota and L.~Musa,
  The ALTRO chip: A 16-channel A/D converter and digital processor for gas 
  detectors,
  IEEE Trans.\ Nucl.\ Sci.\  {\bf 50}, 2460 (2003).
  %doi:10.1109/TNS.2003.820629
\bibitem{RCU}
    C.~G.~Gutierrez et al., The ALICE TPC Readout Control Unit, IEEE Nuclear Science Symposium Conference Record, 1, 575
